The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-09-08 paper/paper.Rnw
Reread and corrected half the paper

\section{Literature Review}
% You have a bunch of citations to take from here
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family SES and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements. For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}. But fast-forward until today, we've understood the relationship much better.

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive and personality definition \citep{duyme1999, waldfogel2006}. Economists and educators have long neglected this branch of literature but the work of James Heckman has brought our attention to it. For example, he has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\). That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns. Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}. What we don't know is how and why this relationship is formed from such an early age. Moreover, many have suggested different explanations and the topic has been very contested \citep{brooks1997}.

It is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that education is by far the investment that yields the best returns. This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities. We still don't know, in comparative terms, which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

Virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitud of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.

From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}. The important finding came when we discovered the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential. It does so very strongly for children which otherwise would be doomed to have a bleak future. Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality. These two traits make the Danish system uniquely effective. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

@@

The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years. The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence \citep{anna2016_global} is emerging with a very similar landscape. The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940. This time-trend has allowed researchers to study achievement gaps in a lengthy period of time. Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}. The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).

@@

However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970 where income inequality was increasing.}.
.

In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend. This follow-up study concentrated solely on the kindergarten children in which they took a snapshot of the achievement gap in for 1998, 2006 and 2010. They found that the 90th/10th income gap in readiness closed modestly. Furthermore, using data from fall and spring in the same kindergarden year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010. They also calculated the same changes for a number of personality traits such as self-control and externalizing behaviour and found similar results. In contrast to \citet{reardon2011}, he finds that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates. Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap). It should be noted that perhaps the reversal of trend in \citet{reardon2011} would be obvious if data were available for years after 2000,  which is specifically the time that \citet{reardon_portilla} find the reversal.

The reasons the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to try and explaind this change and suggest that the reversal is likely due to the high increase of preschool enrollment. They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
The reasons the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to try explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment. In fact, the point out that in this same period (1998 - 2010) the SES gap in preschool enrollment decreased substantially. Their conclusions are very speculative and have no empirical support, which is why this is still an open question.

@@

Another limitation of their study is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared. They do an amazing work at making the four surveys comparable and their findings are indeed very reliable.  But we should be careful at intepreting these findings causally and should be taken as suggestive at best. For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citet{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in the gap using much more comparable surveys. However it should be noted that these studies tackle a completely different question from the above, namely to study cross-sectional differences between many countries, instead of over-time analysis of student gaps. Nonetheless, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.

\citet{anna2016}, again using PIRLS, TIMSS and PISA, assess whether there are patterns of cross-national variation in the achievement gap. In other words, does the achievement gap differ between countries? Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries. In comparison to the literature on achievement gaps, they find that the U.S has a gap of   1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around   1.25 to   1 standard deviation in the same year-span. However, these numbers vary a lot and carry a great deal of overlapping uncertainty.

@@

\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries. With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap. However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase. This is revealing because it doesn't really pay off to look at a general average once each country has their own distinctive gap and evolution. Another important finding is that the achievement gap is clearly no universal and should be studied in context.

One clear limitation of their study (as well as \citet{reardon2011}) is that they adjust for the age of each child in all studies. Although for their modeling purposes this is the correct thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime. However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, these modeling strategies are masking age-specific achievement gaps by controlling for age. We clearly see in \citet{reardon_portilla} that there are age-specific gaps, and they do change at a fast pace in very little time.

In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children. The explanation, although very debated, has been gaining much support in recent years. If we remember correctly, countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}. Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}. Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as eliminating tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.

@@

\section{Research questions}

This study seeks to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries. This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems. In addition, extend the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's a tradeoff between good performance and equality over time separately for each country. The advantages of this study are twofold.

@@

More formally, we're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) the rate at which each country-gap is widening/narrowing and;
c) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
d) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;

@@

a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This shows that it is important to study the changes in the achievement gap over time. This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This shows that it is important to study the changes in the achievement gap over time. This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.

b) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}

c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the performance-equality trade off dilemma of tracking might be much more credible than other explanations.

d) We want to test the notion that better performing countries have a smaller degree of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

\section{Methods}

\subsection{Data}

@@

Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI). This variable attempts to capture the social status of the family, without asking for income information. This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure. This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citet{erikson1979}. It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

Yet one of the most relevant variables for our study is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.

>>>>

\section{Literature Review}
% You have a bunch of citations to take from here
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements. For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}. But fast-forward until today, we've understood the relationship much better.

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}. Economists and educators have long neglected these findings but the work of James Heckman has brought our attention to it. For example, he has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\). That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns. Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that education is by far the investment that yields the best returns. This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities. We have very little evidence, in comparative terms, for which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

However, the closest we have is the vast literature on social mobility. Thanks to these findings, we know that virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitud of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.

From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}. The important finding came when we discovered the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities. The education system is completely subsidized for all children, otherwise giving opportunities to children wouldn't have been able to pay. In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}. Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}. These two traits make the Danish system uniquely effective. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

@@

@@

However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970 where income inequality was increasing.}.

In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend. This follow-up study concentrated solely on the kindergarten children in which they took a snapshot of the achievement gap in for 1998, 2006 and 2010. They found that the 90th/10th income gap in readiness closed modestly. Furthermore, using data from fall and spring in the same kindergarden year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010. They also calculated the same changes for a number of personality traits such as self-control and externalizing behaviour and found similar results. In contrast, \citet{reardon2011} finds that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates. Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap). It should be noted that perhaps the reversal of trend in \citet{reardon2011} would be obvious if data were available for years after 2000,  which is specifically the time that \citet{reardon_portilla} find the reversal.

The reasons why the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment. They build on their previous argument by suggesting that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. Their conclusions, although suggestive, are very speculative and have no empirical support, which is why this is still an open question.

@@

Another limitation of the work of \citet{bradbury2015} is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared. They do an amazing work at making the four surveys comparable and their findings are indeed very reliable.  But we should be careful at intepreting these findings causally and should be taken as suggestive at best. For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citet{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in the gap using much more comparable surveys. However it should be noted that these studies tackle a completely different question from the above, namely to study cross-sectional differences between many countries, instead of over-time analysis of student gaps. Nonetheless, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.

\citet{anna2016}, again using PIRLS, TIMSS and PISA, assess whether there are patterns of cross-national variation in the achievement gap. In other words, does the achievement gap differ between countries? Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries. In comparison to the literature on achievement gaps, they find that the U.S has a gap of 1.20 SD in 2001 which increase to around 1.30 in the year 2006 while Germany has a decreasing gap from around 1.25 to 1 standard deviation in the same year-span. However, these numbers vary a lot and carry a great deal of overlapping uncertainty.

@@

\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries. With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap. However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase. This is revealing because it doesn't really pay off to look at a general average once each country has their own distinctive gap and evolution. This highlights the notion that the increasing achievement gap is clearly not universal and should be studied in context.

One limitation of their study (as well as \citet{reardon2011}) is that they adjust for the age of each child in all studies. Although for their modeling purposes this is the correct thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime. However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, these modeling strategies are masking age-specific achievement gaps by controlling for age. We clearly see in \citet{reardon_portilla} that there are age-specific gaps, and they do change at a fast pace in very little time.

In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children. The explanation, although very debated, has been gaining much support in recent years. If we remember correctly, for countries with high levels of curricular differentiation the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}. Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}. Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as eliminating tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.

@@

\section{Research framework}

\subsection{Research questions}

This study seeks to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries and search for a likely explanation. This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems. In addition, it extends the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's an over-time tradeoff between good performance and equality over time separately for each country. The advantages of this study are threefold.

@@

Third, we attempt to model why the achievement gap is changing over time. In contrast to other studies, this is a novelty given that most research has been done cross-sectionally between countries.

More formally, we're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind;
c) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
d) What explains this cross-country evolution;

@@

a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap over time. We will compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}

b) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.

c) We want to investigate whether better performing countries have lower levels of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

d) Tightly linked to the previous point, we want to test whether several dimensions of the tracking setup explain why the gaps are changing. This will support the previous points in suggesting that the dynamics of the over-time achievement gap are endogenously intertwined with the effect that tracking has on the specific changes of the high and low SES groups.

\section{Hypothesis}

Following the previous questions and the literature on educational inequality and tracking we pose three hypothesis for answering questions (b), (c) and (d) given that question (a) is purely an exploratory exercise.

\begin{itemize}
\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at the rate given that tracking is thought to maximize the learning experience of both groups. For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this doesn't mean that the top is loweing their score but that the bottom is most likely catching up).

\item Hypothesis 2: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality. Countries with high performance should have less inequality and countries with low performance should have greater inequality

\item Hypothesis 3: if any (or both) of the previous hypothesis are confirmed, then tracking should play an important role on the evolution of the achievement gap. We hypothesize that the degree of tracking is tightly related to changes in the gap, and the more tracking, the more inequality. Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labour market opportunities.
\end{itemize}

\section{Methods}
\subsection{Data}

@@

Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI). This variable attempts to capture the social status of the family, without asking for income information. This index variable was developed by \citet{ganzeboom1996} and later refined by \citep{ganzeboom2010} and it attempts to measure occupational status using a continuous measure. This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citep{erikson1979}. It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

Yet one of the most relevant variables for our study is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measures the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.
